Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants

### 2018-06-04 paper/reviewer_response_PNAS.tex

\comment{Here we need to explain that constraint in our experiments and conservation in nature are not expected to be entirely the same.
Conservation in nature results from the convolution of constraint and diversifying selection from immunity, whereas our experiments only look at constraint.
As evidence to explain this point, we can point to studies showing that the head of HA is much more conserved for swine influenza viruses where there is less pressure from immunity.
{\bf Juhye:} To make this more convincing, we should also pick a site in the transmembrane domain that is highly conserved in nature but not under constraint in our experiments, and validate it.}

\comment{JML: OK, can test a couple of these as well. I'd also like to point out (1) that the WSN HA also exhibits some degree of flexibility in amino acids that are tolerated in the TM domain, and (2) that various studies have found that chimeric HA's with transmembrane domains from {\bf completely different proteins} can still maintain HA cell surface expression and fusion activity (Melikyan et al 1999 Mol Biol Cell, Schroth-Diez et al 1998 JVI, Dong et al 1992 JVI, Roth et al 1986 JCB) and that the {\bf length} of the TM domain seems to play a large role in determining HA fusion activity (Armstrong et al 2000 JCB).}

4) Were sequences from passaged viruses (isolated on MDCKs and other cell lines) included in the comparison between in vitro mutagenesis data and natural evolution data? They need to be excluded since their presence might induce a bias towards what is tolerated/preferred in cell culture. Only sequences obtained directly from patient specimens should be used for the comparison.

\response{We thank the reviewer for this good suggestion and now include an analysis of the relationship between mutational effect and maximum frequency for mutations from unpassaged isolates.
Due to the sparseness of unpassaged isolate sequences previous to the last decade, we analyzed unpassaged isolates from the post-Perth/2009 partition of the H3N2 HA phylogenetic tree.
Figures 5, 6, and S6 now include panels showing the results for these unpassaged isolate mutations.
The results in Figure 5 demonstrate that the correlation between mutational effect as determined by our experiments and maximum mutation frequency is stronger for unpassaged isolates than when we include all isolates ($\rho = 0.24$).
We do not see such a correlation when we use the WSN/1933 H1 HA preferences to estimate the mutational effects for unpassaged isolate mutations.
}

@@

\comment{This is a good suggestion. 
{\bf John:} I think we should add an additional analysis that does this only for sequences that were not passaged in cell-culture or eggs.
I believe that GISAID annotates passage history.
I would expect that we can only do this for the post-Perth sequences as I doubt that there are many unpasaged viruses prior to the early 2000's.
We could just add this as another panel assuming it gives the same results.}

5) It is unclear how the ``rescaling'' is performed and what it exactly does. The authors need to described and justify this better and should also provide the non-rescaled data in a ``sequence logo'' figure for comparison.

\response{Rescaling of the preferences is performed by fitting a stringency parameter that is a reflection of how strongly our experimentally determined preferences correspond to the gene's evolution in nature.
This parameter is fit under an experimentally informed codon substitution model (ExpCM) using the phydms program \comment{reference here}.
A stringency parameter greater than 1 indicates that natural selection favors the same amino acids as the experiments but with greater stringency, while a value less than 1 indicates that natural selection does not favor the amino acids with as much stringency as the experiments.
Because selection in the lab often does not fully reflect the stringency of selection in nature and selection across experiments may not be entirely consistent, we re-scale the preferences by this stringency parameter.
We now include a supplementary panel A to Figure S3 showing the un-scaled Perth/2009 H3 HA preferences.
}

@@

The reviewer is correct that G78D appears to be a site where the correspondence between selection in our experiments and frequency in nature is low.
This makes sense, as G78D is one of the two sites that we specifically identified as being a lab adaptation (see Figure~S1).
We now explicitly mention this fact in the text \comment{still need to actually add this mention}.
}

The data regarding greater mutational tolerance in the stalk compared to the head should be paired with data from deep mutational scanning utilizing stalk- and head-specific antibodies as a selection process as this group has previously done with WSN H1 (Doud et al., Nat Comm 2018, Doud et al., PLoS Path 2017) to see if tolerant sites are responsible for viral escape.

\response{We agree that this would be an interesting experiment to perform...}

@@

\comment{We can definitely do this pretty easily, perhaps adding Sarah Hilton as an author to help.
Not sure where we would fit the data given that the paper is already too long, but I guess we could cram it in the supplement.
}

\comment{JML: I've also been wanting to do this and can do this easily. But it depends on how much space we have (probably not much)}.

The authors are upfront about the limited degree to which their calculated mutational effect and the maximum mutation frequency in nature correlate, describing it as ``modest.'' One of the major points of the paper is that conclusions from this type of analysis cannot be generalized when using data from strains that are too divergent. The authors should address the possibility that data from deep mutational scanning can be generalized to divergent strains if the study is done in a system that is more reflective of nature.

\response{We now have acknowledged this possibility in the Discussion in the sentence "We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility and possibly their generalizability to more divergent strains.}

@@

\comment{This is a fair point.
However, I think we should argue that it's the first paper to show deep mutational scanning can actually provide at least some information that reflects outcomes in nature.
This comment does make me worry that the reviewer just doesn't think the work is significant enough, but since the editor did ask for a revision you'd have to hope that if we answer all specific comments that would be enough.
}

\subsubsection*{Minor Points:}

The correlation for prospective analysis (post-Perth) is smaller than the correlation for post-hoc analysis (pre-Perth). Since prospective analysis is more important when predicting evolutionary fates, it might be beneficial for the authors to comment on this difference.

\response{We now include an analysis of mutational effects and maximum frequency for mutations of unpassaged isolates (see Figures 5, 6, and S6) from the post-Perth partition of the phylogenetic tree to limit any potential biases that our experimental measurements may have towards mutations from cell- or egg-passaged isolates.
The results show that this correlation is stronger when considering mutations from unpassaged isolates in a post-hoc analysis, demonstrating the potential utility of our experimental measurements in a prospective analysis, specifically when passaged isolate sequences are excluded.
}

@@

In figure 4, it would be useful to show Perth09 on the tree and mention that closely related strains are excluded from analysis in the figure legend. As it stands, there is just an unlabeled cluster of viruses for which no individual amino acid mutations are characterized. It is also important for the authors to clearly state what frequency they are talking about. Specifically, in consideration of a particular mutation, what is the population in the denominator?

\comment{We can clarify these easily.}

The graph labels in figure 8B could be cleaner. The question marks after "absolutely conserved" and "clade-specific" are extraneous. Indeed, the graph labels themselves ("HA domain," "absolutely conserved? ""clade-specific?") are unnecessary.

\response{We have now fixed the graph labels in figure 8B accordingly.}

### 2018-06-04 paper/manuscript.tex
Start integrating new figures, update paper text, fig captions, start updating reviewer responses

\begin{SCfigure*}[\sidecaptionrelwidth][h!]
\centering
\includegraphics[width=12cm]{figs/muteffect_by_maxfreq/muteffect_by_maxfreq.pdf}
\caption{\label{fig:muteffect_maxfreq}
{\bf Experimental measurements are informative about the evolutionary fate of viral mutations.}
(A) Correlation between the effects of mutations as measured in our deep mutational scanning of the Perth/2009 HA and the maximum frequency reached by these mutations in nature.
The plots show Spearman $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations of the experimental measurements for which the permuted $\rho$ was greater than or equal to the observed $\rho$.
(B) The distribution of mutational effects partitioned by maximum mutation frequency.
The vertical black line shows the mean mutation effect for each category.
The analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
}
\end{SCfigure*}

@@

We next sought to quantify the correlation between a mutation's experimentally measured effect and the maximum frequency it attained during natural evoluton.
To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after that have not had enough time for their evolutionary fates to be fully resolved.
We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
We quantified the relationship between mutational effects and maximum mutation frequencies in the H3N2 phylogeny via Spearman rank correlation (Figure~\ref{fig:muteffect_maxfreq}A).
In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$).
The similar effect sizes for both partitions shows that our experimental measurements can help explain the evolutionary fates of mutations in strains that post-date the experimentally studied strain, as well as to retrospectively analyze mutations that precede the experimental strain.

@@

It is important to emphasize that measurements of viral growth in cell culture do \emph{not} represent true fitness in nature.
Indeed, a vast amount of work in virology has chronicled the ways in which experiments can select for lab artifacts or fail to capture important pressures that are relevant in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}.
Mutations in viral genes other than HA are also important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
Given these caveats, it might seem surprising that measuring viral growth in cell culture can be informative about the success of viral strains in nature.
Yet, prior to our work, there were no comprehensive studies of the functional effects of mutations to H3 HA on any property that even resembled viral fitness in nature, and modeling work has either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or assumed that all non-epitope mutations had equivalent deleterious effects~\cite{luksza2014predictive}.
The strength of our measurements are not that they perfectly capture fitness in nature, but that they are systematic and quantitative---and so represent an improvement over no information at all.
We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility.

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=\textwidth]{figs/supp_WSNprefs_logoplot/WSN-rescaled_prefs.pdf}}
\caption{\label{suppfig:WSNprefs_logoplot}
{\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA as measured in \cite{doud2016accurate}.}
The amino-acid preferences from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
The sites are in H3 numbering.
The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
Note that \cite{doud2016accurate} used libraries in which all codons were mutagenized \emph{except} for the one encoding N-terminal methionine.
Therefore, no data is shown for the first codon in the gene.
}
\end{suppfigure}

\begin{suppfigure}[H]
\centerline{\includegraphics[width=1.0\textwidth]{figs/supp_H3N2_phylogeny/H3N2_phylogeny.pdf}}
\caption{\label{suppfig:tree}
{\bf A phylogenetic tree of all HA sequences used in our analysis of mutation frequencies.}
HA sequences were sampled at a rate of six viruses per month from January 1, 1968 through February 1, 2018.
The Perth/2009 strain used in our experiments is indicated.
The rest of the tree is partitioned into nodes that preceded the split of the Perth/2009 strain from the trunk of the tree (blue) and nodes that branched off the trunk after the clade containing Perth/2009 (orange).
In Figure~\ref{fig:muteffect_maxfreq}, these two partitions of the tree are analyzed separately.
Nodes in the clade containing the Perth/2009 strain and nodes sampled in 2014 or after were excluded from our analyses.
The Perth/2009 strain was excluded to avoid artifacts related to mutations that occurred on the branches leading to the HA sequence used in the experiment.
The post-2014 nodes were excluded because the evolutionary fates of many sequences after this date are not yet full resolved.
}
\end{suppfigure}

>>>>

\begin{SCfigure*}[\sidecaptionrelwidth][h!]
\centering
\includegraphics[width=12cm]{figs/muteffect_by_maxfreq/muteffect_by_maxfreq.pdf}
\caption{\label{fig:muteffect_maxfreq}
{\bf Experimental measurements are informative about the evolutionary fate of viral mutations.}
(A) Correlation between the effects of mutations as measured in our deep mutational scanning of the Perth/2009 HA and the maximum frequency reached by these mutations in nature.
The plots show Spearman $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations of the experimental measurements for which the permuted $\rho$ was greater than or equal to the observed $\rho$.
(B) The distribution of mutational effects partitioned by maximum mutation frequency.
The vertical black line shows the mean mutation effect for each category.
The analysis is performed separately for pre-Perth/2009, post-Perth/2009, and unpassaged isolates from the post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
}
\end{SCfigure*}

@@

We next sought to quantify the correlation between a mutation's experimentally measured effect and the maximum frequency it attained during natural evolution.
To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after that have not had enough time for their evolutionary fates to be fully resolved.
We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
We quantified the relationship between mutational effects and maximum mutation frequencies in the H3N2 phylogeny via Spearman rank correlation (Figure~\ref{fig:muteffect_maxfreq}A).
In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$). 
To restrict potential biases of our experimental measurements towards mutations from cell- or egg-passaged isolates, we evaluated mutations from unpassaged isolates.
Due to the limited availability of primary isolate sequences until the last decade, we analyzed mutations of unpassaged isolates from the post-Perth partition of the phylogenetic tree~\cite{mcwhite2016sequence}.
The correlation between mutational effect and maximum frequency was stronger for mutations from unpassaged post-Perth isolates ($\rho = 0.24$).
The similar effect sizes for both partitions shows that our experimental measurements can help explain the evolutionary fates of mutations in strains that post-date the experimentally studied strain, as well as to retrospectively analyze mutations that precede the experimental strain.

@@

It is important to emphasize that measurements of viral growth in cell culture do \emph{not} represent true fitness in nature.
Indeed, a vast amount of work in virology has chronicled the ways in which experiments can select for lab artifacts or fail to capture important pressures that are relevant in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}.
Mutations in viral genes other than HA are also important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
Given these caveats, it might seem surprising that measuring viral growth in cell culture can be informative about the success of viral strains in nature.
Yet, prior to our work, there were no comprehensive studies of the functional effects of mutations to H3 HA on any property that even resembled viral fitness in nature, and modeling work has either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or assumed that all non-epitope mutations had equivalent deleterious effects~\cite{luksza2014predictive}.
The strength of our measurements are not that they perfectly capture fitness in nature, but that they are systematic and quantitative---and so represent an improvement over no information at all.
We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility and possibly their generalizability to more divergent strains.

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_WSNprefs_logoplot/supp_logoplots.pdf}}
\caption{\label{suppfig:WSNprefs_logoplot}
{\bf The site-specific amino-acid preferences of the Perth/2009 H3 HA and the WSN/1933 H1 HA.}
(A) The un-scaled amino-acid preferences of the Perth/2009 H3 HA.
The overlays show the same information as in Figure~\ref{fig:logoplot} (domain, epitope site, and wildtype amino acid).
(B) The amino-acid preferences of the WSN/1933 H1 HA from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
Note that \cite{doud2016accurate} used libraries in which all codons were mutagenized \emph{except} for the one encoding N-terminal methionine.
Therefore, no data is shown for the first codon in the gene.
The sites are in H3 numbering for both logo plots.
}
\end{suppfigure}

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.9\textwidth]{figs/supp_H3N2_phylogeny/H3N2_phylogeny.pdf}}
\caption{\label{suppfig:tree}
{\bf A phylogenetic tree of all HA sequences used in our analysis of mutation frequencies.}
(A) HA sequences were sampled at a rate of six viruses per month from January 1, 1968 through February 1, 2018.
The Perth/2009 strain used in our experiments is indicated.
The rest of the tree is partitioned into nodes that preceded the split of the Perth/2009 strain from the trunk of the tree (blue) and nodes that branched off the trunk after the clade containing Perth/2009 (orange).
In Figure~\ref{fig:muteffect_maxfreq}, these two partitions of the tree are analyzed separately.
Nodes in the clade containing the Perth/2009 strain and nodes sampled in 2014 or after were excluded from our analyses.
The Perth/2009 strain was excluded to avoid artifacts related to mutations that occurred on the branches leading to the HA sequence used in the experiment.
The post-2014 nodes were excluded because the evolutionary fates of many sequences after this date are not yet full resolved.
(B) The post-Perth/2009 partition of the tree containing only sequences from unpassaged isolates.
}
\end{suppfigure}
